Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-01-30 PAPER.tex

The major challenge that faces all 21\,cm experiments is isolating a small signal that is buried underneath foregrounds and 
instrumental systematics that are, when combined, four to five orders of magnitude brighter \citep[e.g.,][]{santos_et_al2005, ali_et_al2008, deOliveiraCosta_et_al2008, jelic_et_al2008, bernardi_et_al2009, bernardi_et_al2010, ghosh_et_al2011, pober_et_al2013, dillon_et_al2014, kohn_et_al2016}. A clean measurement therefore requires an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made 
in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact 
with each other to affect power spectrum results. More specifically, it is imperative to develop techniques and tests that ensure 
the accuracy and reliability of a potential EoR detection. In this paper, we discuss three topics (signal loss, errors, and bias) that are essential to investigate 
for a robust $21$ cm power spectrum analysis. We also highlight four power spectrum techniques and their trade-offs, potential 
pitfalls, and connections to the themes. We first approach the themes from a broad perspective, and then perform a detailed 
case study using data from the 64-element configuration of PAPER, motivating a revised PAPER-64 power spectrum 
from the lessons learned.

This paper adds to the growing foundations of lessons which have been documented in  \cite{Patil2016,Jacobs2016} and \cite{Paciga2013} by the LOFAR, MWA and GMRT projects respectively.

@@

In Section \ref{sec:SiglossOverview}, we showed how signal loss arises when weighting data using information from the data 
itself. Here we describe a methodology for estimating the amount of signal loss caused by a particular power spectrum estimator when applied to a particular data set. The exact amount of signal loss for a particular data set will depend on the specific realizations of the signals present in the data and is not something we can directly compute. 
In this work, as in A15, we inject simulated cosmological signals into our data and test the recovery of those signals. As we will see, correlations between the injected signals and the data are significant complicating factor which was previously not taken into account.

The results of such an injection scheme can be interpreted in multiple ways.  In this paper we describe two. This section describes one and in Appendix \ref{sec:Pr_appendix} we present an alternate interpretation.
These two methods yield very similar results, but fundamentally are answering subtly distinct questions.  
We also highlight major differences from the signal loss 
computation used in A15, which previously underestimated losses.

@@

The correlation between $\widehat{\textbf{C}}_r^{-1}\textbf{e}$ and $\widehat{\textbf{C}}_r^{-1}\textbf{x}$ means that we cannot compute signal loss using a signal-only 
simulation, which would yield greater values for $P_{\rm out}$ and thereby underestimate signal loss. Therefore, in our revised 
signal loss computation we use the full quantity for $P_{\rm out}$ as defined in Equation \eqref{eq:sigloss}, despite the fact that the cross-terms muddy the waters.

@@

We saw in Section \ref{sec:otherweight} how limiting the number of down-weighted eigenmodes (i.e. flattening out part of the 
eigenspectrum and effectively de-coupling the `weak' eigenmodes from the data) can help minimize signal loss. We experiment with this idea on PAPER-64 data, dialing the number of modes 
that are down-weighted from zero (which is equivalent to identity-weighting, or the unweighted case) to $21$ (which is full inverse 
covariance weighting of our $21$ channels). The power spectrum results for one $k$ value, both before and after signal loss 
estimation, are shown in the top panel in Figure \ref{fig:sigloss_modeloop}. We see that the amount of signal loss increases as weighting 
becomes more aggressive (gray curve). In other words, more `weak' (EoR-dominated) fluctuations are being overfit and 
subtracted as more modes are down-weighted. We also find that the power spectrum upper limit, post signal loss estimation, 
increases with the number of down-weighted modes (black curve). The more modes we use in down-weighting, the stronger the coupling between the weighting and the data, and the greater the error we have in estimating the power spectrum.

Optimistically, we expect there to be a `sweet spot' as we dial our regularization knob; a level of regularization where weighting 
is beneficial compared to not weighting (blue dashed line). In other words, we would like a weighting scheme that down-weights eigenmodes that predominantly describe foreground modes, but not EoR modes. We see in Figure \ref{fig:sigloss_modeloop} that this occurs when 
only the strongest $\sim3$-$4$ eigenmodes are down-weighted, though the improvement from the unweighted case is small.

@@

\begin{itemize}
\item Original Data (black): This is identical to the unweighted, revised PAPER-64 power spectrum in Figure \ref{fig:ps1_data} (one baseline type only) with weighting matrix $\textbf{I}$. There are clear detections at 
low $k$'s.
\item Even/Odd (blue): We split our dataset into even (\textbf{e}) and odd (\textbf{o}) days. We then form the following two datasets: $\textbf{x}_{1} = \textbf{e} + \textbf{o}$ and $\textbf{x}_{2} = 
\textbf{e} - \textbf{o}$. We form datasets in this way to ensure that we use the full sensitivity of our data. When cross-multiplied, we obtain:
\begin{equation}
\textbf{x}_{1}\textbf{x}_{2}^{\dagger} = \textbf{ee}^{\dagger} - \textbf{eo}^{\dagger} + \textbf{oe}^{\dagger} - \textbf{oo}^{\dagger}
\end{equation}
If the same sky signal is in both the even and odd datasets, we expect it to cancel out.
\item Baselines (green): We split our dataset into two halves, where each contains half of the total baselines used in the 
analysis. Again, we form $\textbf{x}_{1} = \textbf{b}_{1} + \textbf{b}_{2}$ and $\textbf{x}_{2} = \textbf{b}_{1} - \textbf{b}_{2}$, where $\textbf{b}_{1}$ is the first baseline 
group and $\textbf{b}_{2}$ is the second baseline group.
\item LST (magenta): We split our dataset into two halves based on LST, namely $\textbf{t}_{1}$ (LSTs 0.5-4.6 hours) and $\textbf{t}_{2}$ (LSTs 
4.6-8.6 hours). We form our two datasets as $\textbf{x}_{1} = \textbf{t}_{1} + \textbf{t}_{2}$ and $\textbf{x}_{2} = \textbf{t}_{1} -\textbf{t}_{2}$.
\end{itemize}

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/null.png}
	\caption{Results for three null tests compared to the PAPER-64 unweighted revised power spectrum (black) and analytically estimated 
$2\sigma$ errors (gray shaded region). For each, we take a jackknife along a different axis of the dataset - along julian days (separating 
even and odd days; blue), along baselines (green), and along LST (magenta). We expect the sky signal to disappear for a `passing' null test (for all error bars to be contained to within the gray shaded region).
We find that none of the null tests performed pass, and that the dominant source of the biases in our power spectrum are likely caused by variation of foregrounds in LST (due to the magenta result being the most inconsistent with noise)}
	\label{fig:null}
\end{figure*}

Investigating Figure \ref{fig:null}, we find that none of the null tests are fully consistent with noise; we would expect the null tests' error bars to be contained within the gray shaded region (theoretical error) for a `passable' test. We note, however, that the error bars for the baselines test (green) and even/odd test (blue) are smaller than those of the LST test (magenta).

%We note that the error bars for the baselines test (green) are especially small --- this is because the signal is effectively cancelled out prior to cross-multiplication (when forming baseline groups).

This suggests that the primary systematic driving our power spectrum points and errors to be outside the 
estimated error region is LST-dependent, and therefore likely to be residual foregrounds. In other words, the fact that many magenta power spectrum points are outside the estimated error region implies that $t_{1}$ and 
$t_{2}$ differ by an amount greater than thermal noise and they do not effectively cancel out the sky signal during cross-multiplication. While residual foregrounds are the primary cause for power spectrum results above the theoretical noise (as in Figure \ref{fig:ps1_data}), other unexplained variations exist which are present in all the null tests. These variations could be attributable to... \cc{fill in}

We do not investigate the detailed nature of the residual foregrounds in this paper, though one could imagine performing many 
similar null tests with a sliding window of $t_{1}$'s and $t_{2}$'s. This would illuminate the particular LST range that introduces 
the foreground bias (or whether the bias is constant in LST), and could potentially be traced to an individual bright radio source. 
This type of detailed analysis will be critical at EoR sensitivities; however, for our current analysis we are not surprised that the main source of
bias we see comes from foregrounds in regions where we expect leakage.

>>>>

The major challenge that faces all 21\,cm experiments is isolating a small signal that is buried underneath foregrounds and 
instrumental systematics that are, when combined, four to five orders of magnitude brighter \citep[e.g.,][]{santos_et_al2005, ali_et_al2008, deOliveiraCosta_et_al2008, jelic_et_al2008, bernardi_et_al2009, bernardi_et_al2010, ghosh_et_al2011, pober_et_al2013, bernardi_et_al2013, dillon_et_al2014, kohn_et_al2016}. A clean measurement therefore requires an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made 
in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact 
with each other to affect power spectrum results. More specifically, it is imperative to develop techniques and tests that ensure 
the accuracy and reliability of a potential EoR detection. In this paper, we discuss three topics (signal loss, errors, and bias) that are essential to investigate 
for a robust $21$ cm power spectrum analysis. We also highlight four power spectrum techniques (fringe-rate filtering, weighting, bootstrapping, jackknife testing) and their trade-offs, potential 
pitfalls, and connections to the themes. We first approach the themes from a broad perspective, and then perform a detailed 
case study using data from the 64-element configuration of PAPER, motivating a revised PAPER-64 power spectrum 
from the lessons learned. More specifically, we develop new techniques for computing signal loss and under-go a rigorous re-analysis of PAPER-64 data in order to present a new power spectrum limit. A companion paper, Kolopanis et al., \textit{in prep}, builds off of the methods in this paper to present PAPER-64 results for multi-redshifts and multiple baselines.

Finally, this paper adds to the growing foundations of lessons which have been documented in  \citet{Patil2016}, \citet{Jacobs2016}, and \cite{Paciga2013} by the LOFAR, MWA and GMRT projects respectively. These lessons are imperative as the community as a whole moves towards higher sensitivities and potential EoR detections.

@@

In Section \ref{sec:SiglossOverview}, we showed how signal loss arises when weighting data using information from the data

itself. Here we describe a methodology for estimating the amount of signal loss caused by a particular power spectrum estimator when applied to a particular data set. The exact signal loss ``correction factor" for a particular data set will depend on the specific realizations of the signals present in the data and is not something we can directly compute. 
In this work, we inject simulated cosmological signals into our data and test the recovery of those signals in order to statistically quantify 
the amount of signal loss accompanying a weighting scheme used in OQE power spectrum estimation (an approach also taken by \citet{masui_et_al2013}).  
The results of such an injection scheme can be interpreted in multiple ways.  In this paper we describe two. This section describes one and in Appendix \ref{sec:Pr_appendix} we present an alternate interpretation.
These two methods yield very similar results, but fundamentally are answering subtly distinct questions.  
We also highlight major differences from the signal loss 
computation used in A15, which previously underestimated losses.

@@

The correlation between $\widehat{\textbf{C}}_r^{-1}\textbf{e}$ and $\widehat{\textbf{C}}_r^{-1}\textbf{x}$ means that we cannot compute signal loss using a signal-only 
simulation, which would yield greater values for $P_{\rm out}$ and thereby underestimate signal loss. This argument is also made in \citet{switzer_et_al2015} in regards to estimating loss associated with foreground cleaning of Great Bank Telescope (GBT) data. Therefore, in our revised 
signal loss computation we use the full quantity for $P_{\rm out}$ as defined in Equation \eqref{eq:sigloss}, which subtracts the 
weighted power spectrum of the data from the weighted power spectrum of data plus EoR.

@@

We saw in Section \ref{sec:otherweight} how limiting the number of down-weighted eigenmodes (i.e. flattening out part of the 
eigenspectrum and effectively de-coupling the `weak' eigenmodes from the data) can help minimize signal loss. We experiment with this idea on PAPER-64 data, dialing the number of modes 
that are down-weighted from zero (which is equivalent to identity-weighting, or the unweighted case) to $21$ (which is full inverse 
covariance weighting of our $21$ channels). The power spectrum results for one $k$ value, both before and after signal loss 
estimation, are shown in the top panel in Figure \ref{fig:sigloss_modeloop}. We see that the amount of signal loss increases as weighting 
becomes more aggressive (gray curve). In other words, more `weak' (EoR-dominated) fluctuations are being overfit and 
subtracted as more modes are down-weighted. We also find that the power spectrum upper limit, post signal loss estimation, 
increases with the number of down-weighted modes (black curve). The more modes we use in down-weighting, the stronger the coupling between the weighting and the data, and the greater the error we have in estimating the power spectrum. \citet{switzer_et_al2013} took a similar approach in determining the optimal number of modes to down-weight in GBT data, finding similar trends and noting that removing too few modes is limited by residual foregrounds and removing too many modes is limited by large error bars and signal loss.

Optimistically, we expect there to be a `sweet spot' as we dial our regularization knob; a level of regularization where weighting 
is beneficial compared to not weighting (blue dashed line). In other words, we would like a weighting scheme that down-weights eigenmodes that predominantly describe foreground modes, but not EoR modes. We see in Figure \ref{fig:sigloss_modeloop} that this occurs when 
only the strongest $\sim3$-$4$ eigenmodes are down-weighted and the rest are given equal weights. For a similar discussion on projecting out modes (zero-ing out eigenmodes, rather than just ignoring their relative weightings as we do in this study), see \citet{switzer_et_al2014}.

@@

\begin{itemize}
%\item Original Data (black): This is identical to the unweighted, revised PAPER-64 power spectrum in Figure \ref{fig:ps1_data} (one baseline type only) with weighting matrix $\textbf{I}$. There are clear detections at low $k$'s.
\item Even/Odd (blue): We split our dataset into even (\textbf{e}) and odd (\textbf{o}) days. We then form the following two datasets: $\textbf{x}_{1} = \textbf{e} + \textbf{o}$ and $\textbf{x}_{2} = 
\textbf{e} - \textbf{o}$. We form datasets in this way to ensure that we use the full sensitivity of our data. When cross-multiplied, we obtain:
\begin{equation}
\textbf{x}_{1}\textbf{x}_{2}^{\dagger} = \textbf{ee}^{\dagger} - \textbf{eo}^{\dagger} + \textbf{oe}^{\dagger} - \textbf{oo}^{\dagger}
\end{equation}
If the same sky signal is in both the even and odd datasets, we expect it to cancel out.
\item Baselines (green): We split our dataset into two halves, where each contains half of the total baselines used in the 
analysis. Again, we form $\textbf{x}_{1} = \textbf{b}_{1} + \textbf{b}_{2}$ and $\textbf{x}_{2} = \textbf{b}_{1} - \textbf{b}_{2}$, where $\textbf{b}_{1}$ is the first baseline 
group and $\textbf{b}_{2}$ is the second baseline group.
\item LST (magenta): We split our dataset into two halves based on LST, namely $\textbf{t}_{1}$ (LSTs 0.5-4.6 hours) and $\textbf{t}_{2}$ (LSTs 
4.6-8.6 hours). We form our two datasets as $\textbf{x}_{1} = \textbf{t}_{1} + \textbf{t}_{2}$ and $\textbf{x}_{2} = \textbf{t}_{1} -\textbf{t}_{2}$.
\end{itemize}

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/null_zoom.png}
	\caption{Results for three null tests compared to analytically estimated 
$2\sigma$ errors (gray shaded region). For each of the tests, we take a jackknife along a different axis of the dataset - along julian days (separating 
even and odd days; blue), along baselines (green), and along LST (magenta). We expect the sky signal to disappear for a `passing' null test.
We find that the baselines (green) and even/odd (blue) tests are consistent with noise, and that the dominant source of the biases in our power spectrum are likely caused by variation of foregrounds in LST (due to the magenta result being the most inconsistent with noise)}
	\label{fig:null}
\end{figure*}

Investigating Figure \ref{fig:null}, we find that the baselines test is fully consistent with noise (all error bars pass through zero), and that the even/odd test is consistent with noise given the $2\sigma$ expectations ($95\%$ of the error bars are very close or passing through zero). We additionally note that the error bars for the baselines test (green) are especially small --- this is because the signal is effectively cancelled out prior to cross-multiplication (when forming baseline groups).

The null test in LST is much more illuminating. Because of the large error bars and variation in its result, this suggests that the primary systematic driving our power spectrum points and errors to be outside the 
estimated error region is LST-dependent, and therefore likely to be residual foregrounds. In other words, the fact that several magenta power spectrum points, as well as their error bars, are outside the estimated error region implies that $t_{1}$ and 
$t_{2}$ differ by an amount greater than thermal noise and they do not effectively cancel out the sky signal during cross-multiplication. We also note that the LST null test is biased high, consistent with the interpretation that our positive power spectrum detections in Figure \ref{fig:ps1_data} are those of foregrounds.

We do not investigate the detailed nature of the residual foregrounds in this paper, though one could imagine performing many 
similar null tests with a sliding window of $t_{1}$'s and $t_{2}$'s. This would illuminate the particular LST range that introduces 
the foreground bias (or whether the bias is constant in LST), and could potentially be traced to an individual bright radio source. 
This type of detailed analysis will be critical at EoR sensitivities; however, for our current analysis we are not surprised that the main source of
bias we see comes from foregrounds. % in regions where we expect leakage.
